# Reproducible Research: Peer Assessment 1
Tested under Win 7 - 32 bit

To use the code set an appropriate working directory


## Loading and preprocessing the data
```{r}
if (!file.exists("activity.csv")) {
  url = "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
	download.file(url, destfile = 'activity.zip')
	unzip('activity.zip')
}
```
Open the dataset and recreate time variable from the interval

```{r}
activity_dataset = read.csv("activity.csv")
activity_dataset$interval = as.factor(sprintf("%02d:%02d", activity_dataset$interval%/%100, activity_dataset$interval%%100))
activity_dataset$datetime = strptime(paste(activity_dataset$date,activity_dataset$interval), "%F %H:%M")
str(activity_dataset)
```


## What is mean total number of steps taken per day?
### Histogram of the total number of steps taken each day
To find total number of steps taken each day we aggregate by day

Days where no observations were made (ie all values for step for that data are NA) are skipped entirely in this section

```{r}
daily_aggr_reduced = aggregate(steps ~ date, sum, data = activity_dataset, na.rm = T)
barplot(daily_aggr_reduced$steps, names.arg = daily_aggr_reduced$date, xlab = "date", ylab = "total number of steps", main = "Total Number of Steps Taken Each Day")
```


### Mean and median total number of steps taken per day
Mean total number of steps per day:
```{r}
mean(daily_aggr_reduced$steps)
```
Median:
```{r}
median(daily_aggr_reduced$steps)
```

NOTE: those values would be lower if different form of aggregation was used

For comparison purposes one can consider the following alternative aggregations
```{r}
#Aggregation 1

#aggregate(steps ~ date, sum, data = activity_dataset, na.rm = T, na.action = NULL)

#Aggregation 2 

#tapply(activity_dataset$steps, activity_dataset$date, sum, na.rm = T)
```
In both cases above days which have only NAs would be counted as 0, instead of NA.

Finding mean and median for alternative interpretation is left out of this analysis


## What is the average daily activity pattern?
Here we use intervals to aggregate all non-NA data
```{r}
within_day_aggr_reduced = aggregate(steps ~ interval, mean, data = activity_dataset, na.rm = T)
plot(within_day_aggr_reduced, pch = 20, main = "Average Number of steps taken within 5 minute interval")
lines(within_day_aggr_reduced)
#minor downside of using factors, type = "l" is ignored by the plot command, so separate lines call is required
```

Max number of steps is taken at the following time:
```{r}
within_day_aggr_reduced[which.max(within_day_aggr_reduced$steps),]
```
which is 8.35 AM, with around 200 steps (panic packing for work i presume)
## Imputing missing values
### Total number of missing values
```{r}
na_number = sum(is.na(activity_dataset$steps))
na_number
```
Total number of missing values is `r na_number `


Using mean value for 5-min interval is used as a replacement for NA
```{r}
temp_dataset = merge(activity_dataset, within_day_aggr_reduced, by = "interval")
#sort dataset by the same variables
temp_dataset = temp_dataset[order(temp_dataset$datetime),]
#impute values
activity_dataset$steps[is.na(activity_dataset$steps)] = temp_dataset$steps.y[is.na(activity_dataset$steps)]
rm(temp_dataset)
```

Now plot the total number of steps taken in each day

```{r}
daily_aggr = aggregate(steps ~ date, sum, data = activity_dataset)
barplot(daily_aggr$steps, names.arg = daily_aggr$date, xlab = "date", ylab = "total number of steps", main = "Total Number of Steps Taken Each Day")
```

Major difference observed is that dataset became wider (since we filled in the days which were entirely missed (for example first day)), but the information provided serves no new information, as overall picture remains the same
## Are there differences in activity patterns between weekdays and weekends?
```{r}
activity_dataset$weekdays = weekdays(activity_dataset$datetime)

within_day_aggr_weekdays = aggregate(steps ~ interval, mean, data = activity_dataset[!(activity_dataset$weekdays %in% c("Saturday", "Sunday")),])
within_day_aggr_weekends = aggregate(steps ~ interval, mean, data = activity_dataset[(activity_dataset$weekdays %in% c("Saturday", "Sunday")),])

par(mfrow = c(2, 1))
plot(within_day_aggr_weekdays, pch = 20, main = "Weekdays")
lines(within_day_aggr_weekdays)
plot(within_day_aggr_weekends, pch = 20, main = "Weekends")
lines(within_day_aggr_weekends)
```

As we can see on workdays major activity apears on pre-work, lunch and post-work hours, with the biggest peak in the morning

In case of weekends activities tend to be distributed more uniformly across the day